---
title: "Hamming Ball query example"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

In this notebook, we show how to perform a Hamming Ball query on a synthetic dataset. 

### Setup

Setting up data generation etc code

```{r}
library(tidyverse)
```

Recitulate package is needed for calling python (requires a python installation with numpy and torch)

```{r}
library(reticulate)
use_condaenv("/Users/kmartens/Library/r-miniconda-arm64")

```

Other imports

```{r}
# rarity
source("R/core.R")
# baseline clustering methods
source("R/baselines.R")
# helper functions for plotting
source("R/helpers.R")
```


Code for synthetic data generation

```{r}

rmatnorm <- function(mean_mat, sd_mat){
  eps <- rnorm(nrow(mean_mat) * ncol(mean_mat), mean=mean_mat, sd=sd_mat)
  matrix(eps, nrow(mean_mat), ncol(mean_mat))
}

generate_synthetic_data <- function(marker_mat, n_cells, noise_sd1 = 0.03, noise_sd2 = 0.18){
  Y <- do.call("rbind", lapply(1:nrow(marker_mat), function(i){
    binary_exprs <- t(replicate(n_cells[i], marker_mat[i, ]))
    epsilon <- rnorm(ncol(binary_exprs) * nrow(binary_exprs))
    mean <- 0.05 + 0.5 * binary_exprs
    sd <- noise_sd1 * (binary_exprs == 0) + noise_sd2 * (binary_exprs == 1)
    exprs <- rmatnorm(mean, sd)
    exprs[exprs < 0] <- 0
    exprs[exprs > 1] <- 0.99
    exprs
  }))
  colnames(Y) <- paste0("marker", 1:ncol(Y))
  Y
}

```

Underlying binary marker signatures

```{r}
# generate all combinations of binary ON/OFF states
marker_mat <- as.matrix(expand.grid(replicate(5, 0:1, simplify = FALSE)))

# visualise these binary signatures
colnames(marker_mat) <- paste0("marker", 1:ncol(marker_mat))
pheatmap::pheatmap(marker_mat, cluster_cols=FALSE, cluster_rows=FALSE)

```

Data

```{r}
n_ground_truth_clusters <- nrow(marker_mat)
n_cells <- rpois(n_ground_truth_clusters, lambda = 50)
Y <- generate_synthetic_data(marker_mat, n_cells)
```

Run Rarity clustering on this data

```{r}
rarity_fit <- Rarity_fit(Y, batch_size = 1024L, n_iter=12000L)
```

### Hamming Ball query

Suppose we're interested in finding cells whose expression signatures are "close" to our signature of interest. In this example, in the signature of interest, only the first and the last marker are ON (the rest are OFF). This will be the query signature:

```{r}
query_signature <- c(
  "marker 1" = 1, 
  "marker 2" = 0, 
  "marker 3" = 0,
  "marker 4" = 0,
  "marker 5" = 1
)
```

Function `hamming_ball_query` helps us search for all those cells whose binary signature deviate from a reference cell type by no more than a given number of markers (i.e. their Hamming distance from the reference signature does not exceed a given radius). Let's try out the query with `radius = 1` and `radius = 2`. 

```{r}
source("R/hamming_ball_query.R")

cluster_idx <- hamming_ball_query(query_signature, rarity_fit, radius = 1)
```

Let's visualise those binary signatures from this Hamming Ball (radius = 1) query

```{r}
pheatmap::pheatmap(rarity_fit$unique_profiles[cluster_idx, ], cluster_cols=FALSE, cluster_rows=FALSE, main="Hamming Ball (r=1)")
```

as well as from a Hamming Ball (radius = 2) query

```{r}
cluster_idx <- hamming_ball_query(query_signature, rarity_fit, radius = 2)
pheatmap::pheatmap(rarity_fit$unique_profiles[cluster_idx, ], cluster_cols=FALSE, cluster_rows=FALSE, main="Hamming Ball (r=2)")

```

